---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
**Instructions: **  
Show any code that is needed to:  
1. Load the data (i.e. read.csv())  
2. Process/transform the data (if necessary) into a format suitable for your
analysis  

```{r}
### Loading and preprocessing the data
# Instructions:
# Show any code that is needed to
# 1. Load the data (i.e. `read.csv()`)
# 2. Process/transform the data (if necessary) into a format suitable for your analysis

# For ReadMe:
# User: download activity data either from github repo or
# https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip
# Save file in working directory


# if file is from github, it will have zipped name of 'activity.zip'
# if file is from web, zipped name will be 'repdata%2Fdata%2Factivity.zip'

# See if data file is in working directory.  If not, get it
if(!file.exists("activity.zip")) {
    # see if download data (not from github) is there
    if(!file.exists("repdata%2Fdata%2Factivity.zip")) {
        # Neither file in working directory, download data
        fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
        download.file(fileURL, "activity.zip")
    } else {
        #downloaded file is there, rename
        file.rename("repdata%2Fdata%2Factivity.zip", "activity.zip")
    }
# extract
unzip("activity.zip")
}


# See if data read into R.  If not, read it.
if(!exists('activity')){
    # set up column classes
    classes <- c('numeric', 'Date', 'numeric')
    # read data
    activity <- read.csv("activity.csv", na.strings = "NA", colClasses = classes)
}

# check activity - 17568 observations of 3 variables
```



## What is mean total number of steps taken per day?
**Instructions: **   
For this part of the assignment, you can ignore the missing values in the dataset.  
1. Make a histogram of the total number of steps taken each day  

```{r}
# 1. Make a histogram of the total number of steps taken each day

# First make a dataset of the total steps per day
dailysteps <- aggregate(x=activity$steps, by=list(activity$date), FUN=sum,
                        na.rm=TRUE)
# dailysteps has 61 obs of 2 variables

# Then make a histogram of the total steps each day
hist(dailysteps$x,
     main="Histogram of Total Steps Per Day",
     xlab="Total Steps Per Day",
     ylab = "Frequency",
     border="blue",
     col="green")

```


2. Calculate and report the mean and median total number of steps taken per day  

```{r}
# I interpreted this request as finding the mean and median of the total steps
# a day calculated in part 1.  This continues to explore the question of how
# many steps this subject took each day.

meansteps <- mean(dailysteps$x, na.rm = TRUE)         #9354.23, when rounded
mediansteps <- median(dailysteps$x, na.rm = TRUE)     #10395

# Display in Rmd as
# The mean of the total number of steps taken per day is 
# **`r format(round(meansteps, 2), nsmall = 2)`**. 
# The median of the total number of steps taken per day is 
# **`r format(round(mediansteps, 2), nsmall = 2)`**. 
```

The mean of the total number of steps taken per day is 
**`r format(round(meansteps, 2), nsmall = 2)`**.  
  
The median of the total number of steps taken per day is 
**`r format(round(mediansteps, 2), nsmall = 2)`**.  
  
## What is the average daily activity pattern?
**Instructions: **  
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis)   
and the average number of steps taken, averaged across all days (y-axis)  

```{r}
# Set up data: Find average number of steps for every 5-minute interval
intervalsteps <- aggregate(x=activity$steps, by=list(activity$interval),
                           FUN=mean,na.rm=TRUE)
# intervalsteps has 288 obs of 2 variables

# Plot using Base package
plot(intervalsteps$Group.1,
     intervalsteps$x,
     type="l",
     main="Average Number of Steps Per 5-Minute Interval",
     xlab="5-Minute Interval",
     ylab= "Number of Steps (Average)",
     col="blue",
     lwd=2)
```

2. Which 5-minute interval, on average across all the days in the dataset,   
contains the maximum number of steps?

```{r}
# Use which.max to return the row index of the highest average step count
# Then subset the data frame to return the 5-minute interval for that
# measurement
maxinterval <- intervalsteps[which.max(intervalsteps$x),1]

# Display in Rmd as
# The 5-minute interval that, on average across all the days in the dataset,
# contains the maximum number of steps is
# **`r maxinterval`** with **`r max(intervalsteps$x)`** steps.
```

The 5-minute interval that, on average across all the days in the dataset,  
contains the maximum number of steps is
**`r maxinterval`** with **`r max(intervalsteps$x)`** steps.


## Imputing missing values
**Instructions: **  
Note that there are a number of days/intervals where there are missing values  
(coded as NA). The presence of missing days may introduce bias into some  
calculations or summaries of the data.  
1. Calculate and report the total number of missing values in the dataset (i.e.  
the total number of rows with NAs)  
2. Devise a strategy for ﬁlling in all of the missing values in the dataset.  
The strategy does not need to be sophisticated. For example, you could use the  
mean/median for that day, or the mean for that 5-minute interval, etc.  
3. Create a new dataset that is equal to the original dataset but with the  
missing data ﬁlled in.  
4. Make a histogram of the total number of steps taken each day and Calculate   
and report the mean and median total number of steps taken per day. Do these  
values diﬀer from the estimates from the ﬁrst part of the assignment? What is  
the impact of imputing missing data on the estimates of the total daily number  
of steps?


## Are there differences in activity patterns between weekdays and weekends?
**Instructions: **  
For this part the weekdays() function may be of some help here. Use the dataset  
with the filled-in missing values for this part.  
  
Create a new factor variable in the dataset with two levels -- "weekday" and  
"weekend" indicating whether a given date is a weekday or weekend day.  
  
Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute  
interval (x-axis) and the average number of steps taken, averaged across all  
weekday days or weekend days (y-axis). The plot should look something like the  
following, which was created using simulated data:  
  
Your plot will look different from the one above because you will be using the  
activity monitor data. Note that the above plot was made using the lattice  
system but you can make the same version of the plot using any plotting system  
you choose.